---
title: "PS5: rmarkdown and plotting"
author: "James Lee"
date: "February 19, 2023"
output: html_document
---

# INFO201 PS5: rmarkdown and plotting

## Load an check data

_1. Load tidyverse_
```{r, warning=FALSE}
library(tidyverse)
```
_2. Load data. How many rows/columns do we have?_
```{r, message=FALSE}
gm <- read_delim("data/gapminder.csv.bz2")
```  
  This data set contains `r nrow(gm)` observations and
  `r ncol(gm)` variables.
  
_3. Print a small sample of data. Does it look OK?_
```{r}
gm %>% 
  sample_n(10) %>% 
  knitr::kable()
```
  With in the data set, there are some missing values in each varables, however, 
  the overall data seems adequate.

## Descriptive statistics
_1. How many countries are there in the dataset? Analyze all three: iso3, iso2 and name._
```{r}
gm %>% 
  summarize(iso3_total = n_distinct(iso3), name_total = n_distinct(name), 
            iso2_total = n_distinct(iso2)) %>% 
  knitr::kable()
```
  There are 253 countries in iso3, 250 in name, and 249 in iso2

_2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?_

_(a) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?_
```{r}
gm %>% 
  group_by(iso2) %>% 
  summarize(num = n_distinct(name)) %>% 
  arrange(desc(num))
```
  There are 2 NA in iso2 comparably to name
  
_(b) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?_
```{r}
gm %>% 
  group_by(name) %>% 
  summarize(num = n_distinct(iso3)) %>% 
  arrange(desc(num))
```
```{r}
gm %>% 
  filter(is.na(name)) %>% 
  group_by(name) %>% 
  filter(is.na(iso2)) %>% 
  distinct(iso3)
```
  There are 4 NA in name comparably to iso3 and they are CHANISL, GBM, KOS, and NLD_CURACAO

_3. What is the minimum and maximum year in these data?_

Minimum year:
```{r}
gm %>% 
  select(`time`) %>% 
  filter(!is.na(time)) %>% 
  min()
```
Maximum year:
```{r}
gm %>% 
  select(`time`) %>% 
  filter(!is.na(time)) %>% 
  max()
```
  
## CO2 emissions

_1. How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?_
```{r}
gm %>% 
  group_by(time) %>% 
  select(time, co2, co2_PC) %>% 
  summarize(CO2missings = sum(is.na(co2)), CO2_PCmissings = sum(is.na(co2_PC))) 
```
1960 has the most missing data.

_2. Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see._
```{r}
co2Total <- gm %>% 
  group_by(iso3, time) %>% 
  filter(!is.na(co2), !is.na(name)) %>% 
  mutate(co2Average = mean(co2)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | iso3 == "ARG" | iso3 == "ARE")

ggplot(data = co2Total) + 
  geom_col(aes(x = iso3, y = co2Average)) +
  labs(title = "CO2")
```

Out of 5 countries, USA has the biggest CO2 emission.

_3. Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the
same countries. What does this figure suggest?_
```{r}
co2PCTotal <- gm %>% 
  group_by(iso3, time) %>% 
  filter(!is.na(co2_PC), !is.na(name)) %>% 
  mutate(co2PCAverage = mean(co2_PC)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | iso3 == "ARG" | iso3 == "ARE")

ggplot(data = co2PCTotal) + 
  geom_col(aes(x = iso3, y = co2PCAverage)) +
  labs(title = "CO2 per capita")
```

For CO2_PC, United Arab Emirates has the biggest CO2 emission per capita, 
and USA is the second.

_4. Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see._
```{r}
gm %>% 
  filter(time == '2016') %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC), !is.na(region)) %>%
  summarize(co2PCAverage = round(mean(co2_PC), 1)) %>% 
  knitr::kable()
```

In year 2016, the average CO2 emission per capita in each region was
Africa: 1.2, Americas: 4.8, Asia: 6.5, Europe: 6.6, Oceania: 4.6.

_5. Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016_
```{r}
gm %>% 
  filter(!is.na(co2_PC), !is.na(region)) %>% 
  filter(time == 1960 | time == 2016) %>% 
  group_by(region, time) %>% 
  mutate(avgCO2_PC = mean(co2_PC)) %>% 
  ggplot(aes(x = region, y = avgCO2_PC, fill = factor(time))) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(title = "Average CO2 per capita in 1960 and 2016")
```

_6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent)._

Largest emitters

```{r}
gm %>% 
  filter(time == "2019") %>%
  filter(!is.na(region)) %>% 
  group_by(region) %>% 
  arrange(co2_PC) %>% 
  slice_head(n = 3) %>% 
  select(region, name) %>% 
  knitr::kable()
```

Smallest emitters

```{r}
gm %>% 
  filter(time == "2019") %>% 
  filter(!is.na(region)) %>% 
  group_by(region) %>% 
  arrange(co2_PC) %>% 
  slice_tail(n = 3) %>% 
  select(region, name) %>% 
  knitr::kable()
```



## GDP per capita
_1. Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there._
```{r warning=FALSE}
sp1960 <- gm %>% 
  filter(time == "1960")

ggplot(data = sp1960) +
  geom_point(mapping = aes(x = GDP_PC, y = lifeExpectancy, color = region,
                           size = totalPopulation)) +
  labs(title = "1960 Life Expectancy")
```

Analyzing the graph, Europe countries tend to have high life expectancies and GDP, where
African countries mostly stays low in life expectancies and GDP.

_2. Make a similar plot, but this time use 2019 data only._
```{r warning=FALSE}
sp2019 <- gm %>% 
  filter(time == "2019")

ggplot(data = sp2019) +
  geom_point(mapping = aes(x = GDP_PC, y = lifeExpectancy, color = region,
                           size = totalPopulation)) +
  labs(title = "2019 Life Expectancy")
```

_3. Compare these two plots and comment what do you see. How has world developed
through the last 60 years?_

Most countries have developed in GDP and life expectancies, however, African countries
tend to remain similar from the last 60 years.

_4. Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size._
```{r}
gm %>% 
  group_by(region) %>% 
  filter(time == "1960", !is.na(lifeExpectancy), !is.na(region)) %>% 
  mutate(avgLe = mean(lifeExpectancy)) %>% 
  summarize(averageLifeExpentancy = unique(avgLe)) %>% 
  arrange(desc(averageLifeExpentancy))
```

The result seems reasonable, Europe with highest life expectancy 
and Africa with the lowest life expectancy.

_5. Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute._
```{r message=FALSE}
gm %>% 
  filter(!is.na(region), !is.na(lifeExpectancy), !is.na(time)) %>%
  filter(time %in% c("1960", "2019")) %>%
  group_by(region, time) %>% 
  summarize(average = mean(lifeExpectancy)) %>% 
  mutate(previous = lag(average, default = 0), growth = average - previous) %>% 
  arrange(desc(growth))
```

Asia has the highest growth in life expectancy, where Europe 
had the lowest growth in life expectancy.

_6. Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!_
```{r warning=FALSE}
gm %>% 
  filter(!is.na(GDP_PC), time %in% c("1960", "2019")) %>% 
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
     geom_histogram(alpha = 0.5, position = "dodge", bins = 30) +
  scale_fill_manual(values = c("red", "blue"), labels = 
                      c("1960", "2019")) +
  labs(x = "GDP per capita", y = "Amount", title = "GDP per capita in 1960 and 2019") +
  theme_minimal()
```

_7. What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960._
```{r}
gm %>% 
  filter(time == "1960", !is.na(lifeExpectancy), !is.na(region)) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  summarize(name, rank)
```
```{r}
gm %>% 
  filter(time == "2019", !is.na(lifeExpectancy), !is.na(region)) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  summarize(name, rank)
```

In 2019, the USA's life expectancy rank is placed in 46.
_8. If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960._
```{r}
countries2019 <- gm %>% 
  filter(time == 2019, !is.na(lifeExpectancy), name != "") %>% 
  summarise(num_countries_19 = n())

gm %>% 
  filter(time == "2019", !is.na(lifeExpectancy), !is.na(region)) %>% 
  mutate(relativeRank = rank(-lifeExpectancy) / countries2019) %>% 
  filter(name == "United States of America") %>% 
  summarize(name, relativeRank)
```

Relative rank for the USA in 2019 is 0.378